---
title: "Homework - Group A"
author: "Milton Nicolas Plasencia Palacios, Azad Sadr, Gaia Saveri, Alessandro Scardoni"
date: "Spring 2020"
output: 
  html_document:
    toc: yes
  beamer_presentation:
    highlight: tango
  include: null
  ioslides_presentation:
    highlight: tango
  pdf_document:
    highlight: tango
    keep_tex: yes
    toc: yes
  slide_level: 2
  slidy_presentation:
    fig.height: 3
    fig.width: 4
    highlight: tango
header-includes:
- \usepackage{color}
- \definecolor{Purple}{HTML}{911146}
- \definecolor{Orange}{HTML}{CF4A30}
- \setbeamercolor{alerted text}{fg=Orange}
- \setbeamercolor{frametitle}{bg=Purple}
institute: University of Trieste
graphics: yes
subtitle: Simulation, examples and exercises
fontsize: 10pt
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab exercises

## Exercise 1


```{r Lab_ex1, echo=TRUE}
"R code"
```

## Exercise 2
```{r Lab_ex2, echo=TRUE}
"Example 2"
```

## Exercise 3

Repeat the steps above - write the profile log-likelihood, plot it and find the deviance confidence intervals - considering this time $\gamma$ as a nuisance parameter and $\beta$ as the parameter of interest.

---

We fix $\gamma = \hat{\gamma}$, namely the root of:

$$
\frac{\partial \ell(\gamma,\beta;y)}{\partial\gamma} = \frac{n}{\gamma} - n\log(\beta) + \sum_{i=1}^{n}\log(y_i) - \sum_{i=1}^{n}(\frac{y_i}{\beta})^{\gamma}\log(\frac{y_i}{\beta}) = 0
$$

since it is not possible to express $\gamma$ as function of $\beta$ in a closed form. 

```{r Lab_ex3, echo = TRUE}
y<-c(155.9, 200.2, 143.8, 150.1, 152.1, 142.2, 147, 146, 146, 
     170.3, 148, 118, 144, 97)

n<-length(y)

log_lik_weibull<-function(data, param) {
  -sum(dweibull(data, shape=param[1], scale=param[2], log=TRUE))
}

weib.y.mle<-optim(c(1,1),, fn=log_lik_weibull, hessian=T,
                  method='L-BFGS-B', lower=rep(1e-7,2),
                  upper=rep(Inf,2),data=y)

log_lik_weibull_profile<-function(data, beta) {
  gamma.beta<-uniroot(function(x)n/x+sum(log(y))-n*sum(y^x*log(y))/sum(y^x),
                      c(1e-5,15))$root
  log_lik_weibull(data, c(gamma.beta, beta))
}

log_lik_weibull_profile_v<-Vectorize(log_lik_weibull_profile, 'beta')

plot(function(x) -log_lik_weibull_profile_v(data=y, x)+weib.y.mle$value,
     from=120, to=200, xlab=expression(beta),
     ylab='profile relative log likelihood', ylim=c(-10,0))

conf.level<-0.95
abline(h=-qchisq(conf.level,1)/2,lty='dashed',col=2)

lrt.ci1<-uniroot(function(x) 
                 -log_lik_weibull_profile_v(y, x)
                 +weib.y.mle$value
                 +qchisq(conf.level,1)/2,
                  c(1e+7,weib.y.mle$par[2]))$root

lrt.ci1<-c(lrt.ci1,uniroot(function(x) 
                           -log_lik_weibull_profile_v(y,x)
                           +weib.y.mle$value
                           +qchisq(conf.level,1)/2,
                           c(weib.y.mle$par[2],15))$root)

segments( lrt.ci1[1],-qchisq(conf.level,1)/2, lrt.ci1[1],
          -log_lik_weibull_profile_v(y, lrt.ci1[1]), col="red", lty=2)

segments( lrt.ci1[2],-qchisq(conf.level,1)/2, lrt.ci1[2],
          -log_lik_weibull_profile_v(y, lrt.ci1[2]), col="red", lty=2)

points(lrt.ci1[1], -qchisq(0.95,1)/2, pch=16, col=2, cex=1.5)
points(lrt.ci1[2], -qchisq(0.95,1)/2, pch=16, col=2, cex=1.5)
segments( lrt.ci1[1], -10, lrt.ci1[2],-10, col="red", lty =1, lwd=2)
text(158,-9.6,"95% Deviance CI",col=2, cex=0.7)
```

## Exercise 5

In `sim` in the code above, you find the MCMC output which allows to approximate the posterior distribution of our parameter of interest with $\textbf S$ draws of $\theta$. Please produce an histogram for these random draws $\theta^{(1)},\dots,\theta^{(S)}$, compute the empirical quantiles, and overlap the true posterior distribution. 

---

```{r Lab_ex5, echo=TRUE}
library(rstan)
set.seed(42)
#retrieve sim

model_code<-"
data{
  int N;
  real y[N];
  real<lower=0> sigma;
  real mu;
  real<lower=0> tau;
}
parameters{
  real theta;
}
model{
  target+=normal_lpdf(y|theta, sigma);
  target+=normal_lpdf(theta|mu, tau );
}
"

n<-10 #sample size 
sigma2<-2 #likelihood variance
mu<-7 #prior mean
tau2<-2 #prior variance
theta_sample<-2 #true mean
y<-rnorm(n,theta_sample,sqrt(sigma2))

data<-list(N=n, y=y, sigma=sqrt(sigma2), mu=mu, tau=sqrt(tau2))
fit<-stan(model_code = model_code, data=data, chains=4, iter=2000)

sim <- extract(fit) #mcmc output

#posterior mean
mu_star <- ((1/tau2)*mu+(n/sigma2)*mean(y))/((1/tau2)+(n/sigma2))
#posterior standard deviation
sd_star <- sqrt(1/((1/tau2)+(n/sigma2)))

#stan posterior
hist(sim$theta, breaks=40, xlab=expression(theta), ylab="", main="", probability=TRUE)
#true posterior
curve(dnorm(x, mu_star, sd_star), 
       col="blue", lwd=2, add=T)
#empirical quantiles
q<-quantile(sim$theta)
segments(q,0,q, dnorm(q, mu_star, sd_star), col=2)

legend(3.7, 1, c("Stan posterior", "True posterior", "Empirical quantiles"), 
  c("black", "blue", "red"), lty=c(2,1,1),lwd=c(1,1,1), cex=0.5)
```

## Exercise 6

Launch the following line of R code:

```{r Lab6, echo=TRUE}
posterior<-as.array(fit)
```

Use now the `bayesplot` package. Read the help and produce for this example, using the object `posterior`, the following plots:

* posterior intervals.
* posterior areas.
* marginal posterior distributions for the parameters.

Quickly comment. 

---

The object `posterior` is a 3-D array of MCMC draws, for both the parameters `theta` and `lp__`, as observed from the output of `dimnames(posterior)`. We are interested only in `theta`.  

```{r Lab_ex6_int, echo=TRUE}
library(bayesplot)
#posterior intervals
color_scheme_set("brightblue")
mcmc_intervals(posterior, pars=c("theta"))
```

As reported in the `bayesplot` documentation, this is the plot of uncertainty intervals computed from posterior draws with all chains merged, based on quantiles. The thick segment represent the 50% interval, while the thinner outer lines represent the 90% interval. The circle is the posterior median. Since we know the true value of $\theta$ (`mu_star` ≈ 3.158), we can observe that it falls inside the 90% uncertainty interval. 

```{r Lab_ex6_area, warning = FALSE, echo=TRUE}
#posterior areas
color_scheme_set("teal")
mcmc_areas(posterior, pars = c("theta"), prob=0.8, prob_outer = 0.99, point_est="mean") + ggplot2::labs(
  title = "Posterior distribution",
  subtitle = "with mean and 80% intervals"
) 
```

The function `mcmc_areas` shows the uncertainty intervals as shaded areas under the estimated posterior density curve. Here again we can observe that the true value for $\theta$ falls inside the 80% interval.

```{r Lab_ex6_marginal, echo=TRUE, wrning=FALSE}
#marginal posterior distribution for the parameters
color_scheme_set("viridis")
mcmc_hist(posterior, pars=c("theta"), binwidth=0.05)
```

The function `mcmc_hist` plots the marginal posterior distribution of the parameter of interest, namely $\theta$, combining all the chains.